Bitmap Indexing for Fast Analysis of Large Datasets

### 2016-03-31 ms/ms.tex
Changed font sizes in figures to be uniform
Added references to scientific applications of bitmap indexes

%---------------------------------------------------------------------------------
% ABSTRACT
\begin{abstract}
Current scientific datasets can reach unwieldy sizes that impede analysis and scientific discovery. We present a method for using Morton order bitmap indexes to map the parts of the full domain touched by files within a dataset. This takes advantage of domain partitioning across multiple files and limits the volume of data that must be loaded into memory. This method is tested on mock datasets containing $1024^3$ three-dimensional points split across 512 files according to different partitioning schemes. Using Morton order bitmap indexes for localized partitioning schemes like the Hilbert space-filling curve, the number of files that must be loaded to access data for a sub-section of the domain will scale with the size of the sub-section. This will dramatically decrease the time and memory required for loading data during analysis.
\end{abstract}

@@

%---------------------------------------------------------------------------------
% MORTON INDICES
\subsection{Morton Indices}
Morton order indices map multidimensional data onto a one-dimensional space filling curve \citep{Morton1996}. This is done by breaking up the domain into cells where each cell's position within the $N$-dimensional domain can be described by $N$ integers. The Morton index of the cell is then created by interleaving the bits of the $N$ integers to create a single integer that fully describes the cell's position (see Figure \ref{fig:zorder}). 
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.75\columnwidth,keepaspectratio]{../images/zorder.png}
\caption{Example of 3rd order Morton curve. The bits of the $x$ and $y$ indices are interleaved to generate a single integer fully descriptive of it's location within the two-dimensional domain to within $1/2^{3}$th in each dimension.}
\label{fig:zorder}
\end{center}
\end{figure}
%

@@

@@

Collisions can be eliminated by either increasing the order of the index or nesting a second index within cells that contain collisions. This is demonstrated in panel (b) of Figure \ref{fig:collision}. In those cells which contained collisions, a 2nd order Morton index was added. Those cells with collisions at the level of the refined index (purple cells in panel (b)) cover a much smaller portion of the domain than the cells with collisions at the level of the coarse index (purple cells in panel (a)). This means that any given selection is less likely to contain a collision and it will be less likely for a selector to require both files to be loaded unless it actually touches data from both files.

@@

%---------------------------------------------------------------------------------
% COMMENTS
\newpage
\todo{To do:
\begin{itemize}
\item Improve introduction \& conclusion
\item Find \& add tagged references.
\item Acknowledgements
\item Switch times for tests to be averages over 10 runs (with std. dev. error bars?)
\item Make diagram showing buffer added to test datasets?
\item Plot \% files identified vs. \% domain selected?
\item Ensure plot labels are all the same size when in document (set widths to colsize)
\end{itemize}
}

>>>>

%---------------------------------------------------------------------------------
% ABSTRACT
\begin{abstract}
Current scientific datasets can reach unwieldy sizes that impede analysis and scientific discovery. We present a method for using Morton order bitmap indexes to map the parts of the full domain touched by files within a dataset. This takes advantage of domain partitioning across multiple files and limits the volume of data that must be loaded into memory for a single query. This method is tested on mock datasets containing $1024^3$ three-dimensional points split across 512 files according to different partitioning schemes. Using Morton order bitmap indexes for localized partitioning schemes like the Hilbert space-filling curve, the number of files that must be loaded to access data for a sub-section of the domain will scale with the size of the sub-section and can be identified in under 1\,s even when the entire domain is included. This will dramatically decrease the time and memory required for loading data during analysis.
\end{abstract}

@@

%---------------------------------------------------------------------------------
% MORTON INDICES
\subsection{Morton Indices}
Morton order indices map multidimensional data onto a one-dimensional space filling curve \citep{Morton1996}. This is done by breaking up the domain into cells where each cell's position within the $N$-dimensional domain can be described by $N$ integers. The Morton index of the cell is then created by interleaving the bits of the $N$ integers to create a single integer that fully describes the cell's position (see Figure \ref{fig:zorder}). 
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/zorder.png}
\caption{Example of 3rd order Morton curve. The bits of the $x$ and $y$ indices are interleaved to generate a single integer fully descriptive of it's location within the two-dimensional domain to within $1/2^{3}$th in each dimension.}
\label{fig:zorder}
\end{center}
\end{figure}
%

@@

Bitmap indexes are popular for use with large data warehouses, but have also been applied to scientific data from a diverse array of fields including geosciences \citep{Malensek2014}, earth sciences, rocket science \citep{Sinha2006,Sinha2007}, and high-energy physics \citep{Stockinger2000}. \todo{more?}

@@

Collisions can be limited by either increasing the order of the index or allowing for multi-resolution indexes \citep{Sinha2006,Sinha2007}. Panel (b) of Figure \ref{fig:collision} demonstrates an example of nesting a second index within cells that contain collisions. In those cells which contained collisions, a 2nd order Morton index was added. Those cells with collisions at the level of the refined index (purple cells in panel (b)) cover a much smaller portion of the domain than the cells with collisions at the level of the coarse index (purple cells in panel (a)). This means that any given selection is less likely to contain a collision and it will be less likely for a selector to require both files to be loaded unless it actually touches data from both files.

@@

%---------------------------------------------------------------------------------
% COMMENTS
\newpage
\todo{To do:
\begin{itemize}
\item Improve introduction \& conclusion
\item Find \& add tagged references.
\item Acknowledgements
\item Switch times for tests to be averages over 10 runs (with std. dev. error bars?)
\item Make diagram showing buffer added to test datasets?
\item Plot \% files identified vs. \% domain selected?
\item Move figures to the end for submission
\end{itemize}
}
